---
title: "Final-Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyverse)
library(rvest)
library(lubridate)
```

# DS2020: Final Project

Members: Samara Feldhacker (samara28), Alina Kirkpatrick (alinayjk)

### Data Gathering

In order to gather the data from online resources, I created functions that would take the base url and gather all of the information from the website that was needed. (More text yapping)

Creating functions:

```{r}
base_address <- "https://results.truetimeracing.com/"

get_distance <- function(race_name) {
  if(race_name == "4 Mile Run") {
    return(4.0)
  }
  if(race_name == "8 Mile Run") {
    return(8.0)
  }
  if (race_name == "Half Marathon") {
    return(13.1)
  }
  if (race_name == "10k Run") {
    return(6.2)
  }
  if (race_name == "5k Run") {
    return(3.1)
  }
  return(0.0)
}

information_scrapper <- function(url) {
  #First open the base url
  site <- read_html(url)
  
  #First gather all of the races
  race_links <- site %>% html_elements("#ctl00_Content_Main_divEvents .nav-link") %>% html_attr(name="href")
  race_names <- site %>% html_elements("#ctl00_Content_Main_divEvents .nav-link") %>% html_text()
  
  #Get the other page links for that race
  page_links <- site %>% html_elements("#ctl00_Content_Main_grdTopPager a") %>% html_attr(name="href")
  
  #Get the first table
  tables <- site %>% html_table(fill=TRUE)
  data <- tables[[2]]
  colnames(data) <- data[1,]
  data <- data[-1,]
  
  #If there is not distance in col names
  if(!("Dist" %in% names(data))) {
    #Create a new column called distance
    data$Dist <- get_distance(race_names[[1]])
  }
  
  first_iteration = 1
  
  #Iterate through all of the races
  for(race in race_links) {
    i <- 1
    if(!first_iteration) {
      #Read that specific race page
      url <- paste(base_address, race, sep="")
      html <- read_html(url)
      
      #Get the other page links for that race
      page_links <- html %>% html_elements("#ctl00_Content_Main_grdTopPager a") %>% html_attr(name="href")
      
      #Get the first table
      tables <- html %>% html_table(fill=TRUE)
      
      data1 <- tryCatch(tables[[2]], error = function(e) NULL)
      
      if(is.null(data1)) {
        data <- data[, !duplicated(colnames(data))]
        data <- data |>
          filter(Pos != "") |>
          select(-any_of(c("Fav", "Share", "Cat Pos(Gen Pos)")))
        return(data)
      }
      
      colnames(data1) <- data1[1,]
      data1 <- data1[-1,]
      #If there is not distance in col names
      if(!("Dist" %in% names(data1))) {
        #Create a new column called distance
        data1$Dist <- get_distance(race)
      }
      
      #Only rbind with common columns
      common <- intersect(colnames(data), colnames(data1))
      data <- rbind(data[common], data1[common])
    }
    #Iterate through the links in the race
    for(link in page_links) {
      if(!is.na(link)) {
        url2 <- paste(base_address, link, sep="")
        html2 <- read_html(url2)
        
        tables2 <- html2 %>% html_table(fill=TRUE)
        data2 <- tables2[[2]]
        colnames(data2) <- data2[1,]
        data2 <- data2[-1,]
        
        #If there is not distance in col names
        if(!("Dist" %in% names(data2))) {
          #Create a new column called distance
          data2$Dist <- get_distance(race_names[[i]])
        }
        
        #Only rbind with common columns
        common <- intersect(colnames(data), colnames(data2))
        data <- rbind(data[common], data2[common])
      }
    }
    first_iteration <- 0
    i <- i+1
  }
  
  #Filter not needed columns
  data <- data[, !duplicated(colnames(data))]
  
  # Only include all ones where the person finished
  data <- data |>
    filter(Pos != "") |>
    select(!c("Fav", "Share", "Cat Pos(Gen Pos)"))
  
  
  return(data)
}
```

Now I can use these functions in order to gather the tables from the links:

```{r}
center_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1441")
sugar_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1446")
ledges_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1478")
summer_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1492")
yellow_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1575")
macbride_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1593&EId=2")
annett_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1546&EId=1")
jester_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1582")
```

These will be the base data sets that will later get cleaned and combined to make them easier to work with.

...

Issues I had, and the solutions I produced:

...

### Cleaning the data

Now, using the tables produced, I decided to also create one large data set containing all of the races during the 2025 year. I am deciding to match the large data set with the same number of variables as the Center trails and sugar trails data set, but I am choosing to remove Race No, Finish (same as time), and Pace (which can be manual calculated in a much better format).

Step 1: Ensure all dataframes have the same number/type of variables:

```{r}
#Center and sugar are good

#Ledges is missing City, State, Pace, Start, Finish
ledges_trails <- ledges_trails |>
  mutate(City = "", State = "", Pace = "", Start = "", Finish = "")

#Summer is missing City, State, Pace
summer_trails <- summer_trails |>
  mutate(City = "", State = "", Pace = "")

#Yellow trails is missing city, State
yellow_trails <- yellow_trails |>
  mutate(City = "", State = "")

#Macbride has extra Age variable
macbride_trails <- macbride_trails |>
  select(!c("Age"))

#Annett is missing city, State, Pace
annett_trails <- annett_trails |>
  mutate(City = "", State = "", Pace = "")

#Jester is missing city, State
jester_trails <- jester_trails |>
  mutate(City = "", State = "")
```

Step 2: Clean the individual data sets, because I wanted to use another function.

```{r}
data_cleaner <- function(data, trailName) {
  #Remove Race No, Pace, Finish
  dataNew <- data |>
    select(!c("Race No", "Pace", "Finish"))
  
  #Turn gender into a factor
  dataNew$Gender <- factor(dataNew$Gender)
  
  #Turn Pos, Cat Pos, Gen pos, and Dist into a numerical variable
  dataNew$Pos <- as.numeric(dataNew$Pos)
  dataNew$`Cat Pos` <- as.numeric(dataNew$`Cat Pos`)
  dataNew$`Gen Pos` <- as.numeric(dataNew$`Gen Pos`)
  dataNew$Dist <- as.numeric(dataNew$Dist)
  
  #convert time, and start time into the time variable type
  dataNew$Start <- hms(dataNew$Start)
  dataNew$Time <- hms(dataNew$Time)
  
  #Turn category into a factor
  dataNew$Category <- factor(dataNew$Category)
  
  #Turn city and State into a factor - note may need to change later but whatever
  dataNew$City <- factor(dataNew$City)
  dataNew$State <- factor(dataNew$State)
  
  #Add on column for the trailName, and set as factor
  dataNew$Trail <- trailName
  
  dataNew$Trail <- factor(dataNew$Trail)
  
  return(dataNew)
}
```

```{r}
center_trails <- data_cleaner(center_trails, "Center")
sugar_trails <- data_cleaner(sugar_trails, "Sugar")
ledges_trails <- data_cleaner(ledges_trails, "Ledges")
summer_trails <- data_cleaner(summer_trails, "Summer")
yellow_trails <- data_cleaner(yellow_trails, "Yellow")
macbride_trails <- data_cleaner(macbride_trails, "MacBride")
annett_trails <- data_cleaner(annett_trails, "Annett")
jester_trails <- data_cleaner(jester_trails, "Jester")
```

Now, the dataframes will easily combine together into 1

```{r}
trail_races <- rbind(center_trails, sugar_trails, ledges_trails, summer_trails, yellow_trails, macbride_trails, annett_trails, jester_trails)

#Check the levels of Category, and manually rename to make consistent
levels(trail_races$Category)

levels(trail_races$Category) <- c("0-14", "20-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "Overall Winners", "15-19", "Unkown", "15-19", "20-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "70+")
```

### Basic summaries

Include basic summaries if needed

```{r}
summary(trail_races)
```

### Research Questions Solutions

(Questions copied from Final Project Idea)

1)  Which cities/regions are most represented, and do runners from closer locations perform differently than those who travel farther?

...

2)  How does age group affect pace, and does that relationship change across different race distances?

...

3)  How does the performance gap between male and female athletes (measured by fastest race times) vary across age categories?

For easier time comparisons, I made a new column where the time is mutated to be in seconds:

```{r}
Q3_df <- trail_races %>%
  mutate(time_seconds = as.numeric(Time))

Q3_df
```

Now, I will remove the 0.0 distance:

```{r}
Q3_df <- Q3_df %>%
  filter(!is.na(Dist), Dist != 0.0)

unique(Q3_df$Dist) #Checking it was removed (it was)
```

I will also create a new column that removes the category variable "Overall Winners" for simplicity:

```{r}
Q3_df <- Q3_df %>%
  filter(Category != "Overall Winners")

Q3_df # checking it worked
```

Because there are inconsistencies in some of the age catergories across race distances, we will only be removing the following age categories: 0-14, 15-19, 65-69, 70-74, 70+

```{r}
Q3_df <- Q3_df %>%
  filter(!Category %in% c("0-14", "15-19", "65-69", "70-74", "70+"))

unique(Q3_df$Category)

```

I will now find the top 3 runners for each trail, for each distance on the trail, based on gender, and age:

```{r}
top3_runners <- Q3_df %>%
  group_by(Trail, Dist, Category, Gender) %>%
  arrange(time_seconds, .by_group = TRUE) %>%
  mutate(rank_in_group = row_number()) %>%
  filter(rank_in_group <= 3)

top3_runners <- top3_runners %>%
  select(
    Trail,
    Dist, 
    Category,
    Gender,
    Name,
    time_seconds,
    rank_in_group
  )

top3_runners
# Making sure it made a new df instead of altering the original
trail_races
```

I am now going to summarize the average times for each age group and gender per race and distance:

```{r}
gap_run_time <- Q3_df %>%
  group_by(Trail, Dist, Category, Gender) %>%
  summarise(
    avg_time = mean(time_seconds, na.rm = TRUE),
    .groups = "drop"
  )

gap_run_time
```

From here, I will create a grouped bar-chart to visualize the data:

```{r}
#for better labeling
gap_run_time$Dist_label <- paste0("Dist - ", gap_run_time$Dist)


#the actual graph
ggplot(gap_run_time, aes(x = Category, y = avg_time, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  
  scale_fill_manual(values = c("Female" = "pink",
                               "Male" = "lightblue")) +
  
  facet_wrap(~Dist_label) +
  
  labs(
    title = "Performance Gap Between Male and Female Runners",
    x = "Age Category",
    y = "Average Time (seconds)"
  ) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )

```

Conclusion: from this bar chart we can see that male runners consistently have faster average race distances. The performance gap is present in every subgroup but does vary in magnitude depending age. It tends to be smaller in younger age groups and become varied in older age groups, however female average times increase more notably.

Overall, while gender does have an association with performance differences, the size gap is influenced more by age, and varied across race distances.

4)  Is there a relationship between race distance and the number of participants in running events (does race distance affect number of participants)?

First, I need to find the number of participants at the races, keeping in mind I do not want to count those at 0.0, and keeping them grouped by distance as well:

```{r}
participants <- trail_races %>%
  filter(!is.na(Dist), Dist != 0.0) %>%
  group_by(Trail, Dist) %>%
  summarise(
    n_participants = n(),
    .groups = "drop"
  )


participants
```

Now that I have all the participant counts, we can visualize the data. I added red man dots to visualize the mean number of runners in each race distance.

```{r}
ggplot(participants, aes(x = factor(Dist), y = n_participants)) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "red") +
  labs(
    title = "Participants by Race Distance",
    x = "Distance",
    y = "Number of Participants"
  ) +
  theme_minimal()
```

From here, I am going to use descriptive statistics in order to find if there is evidence that race distance affects number of participants.

```{r}
participants_summary <- participants %>%
  group_by(Dist) %>%
  summarise(
    mean_participants = mean(n_participants),
    median_participants = median(n_participants),
    sd_participants = sd(n_participants),
    min_participants = min(n_participants),
    max_participants = max(n_participants),
    .groups = "drop"
  )

cor(participants$Dist, participants$n_participants)
#-0.3384045

participants_summary
```

Conclusion: From my corr analysis, the number we are given is -0.3384045. This means that as there is a weak to moderate negative relationship between race distance and number of participants. This suggests that shorter races tend to attract more runner, while longer races tend to have fewer participants. However, the relationship is not strong, meaning that distance alone does not majorly affect number of participants. The summary statistics and the scatter plot support this, particularly the scatter plat as there is no sense of linearity in it.

Therefore I have come to the conclusion that there is *not a strong* relationship between race distance and number of participants.
