---
title: "Final-Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyverse)
library(rvest)
```

# DS2020: Final Project

Members: Samara Feldhacker (samara28), Alina Kirkpatrick (alinayjk)

### Data Gathering

In order to gather the data from online resources, I created functions that would take the base url and gather all of the information from the website that was needed. (More text yapping)

Creating functions:

```{r}
base_address <- "https://results.truetimeracing.com/"

get_distance <- function(race_name) {
  if(race_name == "4 Mile Run") {
    return(4.0)
  }
  if(race_name == "8 Mile Run") {
    return(8.0)
  }
  if (race_name == "Half Marathon") {
    return(13.1)
  }
  if (race_name == "10k Run") {
    return(6.2)
  }
  if (race_name == "5k Run") {
    return(3.1)
  }
  return(0.0)
}

information_scrapper <- function(url) {
  #First open the base url
  site <- read_html(url)
  
  #First gather all of the races
  race_links <- site %>% html_elements("#ctl00_Content_Main_divEvents .nav-link") %>% html_attr(name="href")
  race_names <- site %>% html_elements("#ctl00_Content_Main_divEvents .nav-link") %>% html_text()
  
  #Get the other page links for that race
  page_links <- site %>% html_elements("#ctl00_Content_Main_grdTopPager a") %>% html_attr(name="href")
  
  #Get the first table
  tables <- site %>% html_table(fill=TRUE)
  data <- tables[[2]]
  colnames(data) <- data[1,]
  data <- data[-1,]
  
  #If there is not distance in col names
  if(!("Dist" %in% names(data))) {
    #Create a new column called distance
    data$Dist <- get_distance(race_names[[1]])
  }
  
  first_iteration = 1
  
  #Iterate through all of the races
  for(race in race_links) {
    i <- 1
    if(!first_iteration) {
      #Read that specific race page
      url <- paste(base_address, race, sep="")
      html <- read_html(url)
      
      #Get the other page links for that race
      page_links <- html %>% html_elements("#ctl00_Content_Main_grdTopPager a") %>% html_attr(name="href")
      
      #Get the first table
      tables <- html %>% html_table(fill=TRUE)
      
      data1 <- tryCatch(tables[[2]], error = function(e) NULL)
      
      if(is.null(data1)) {
        data <- data[, !duplicated(colnames(data))]
        data <- data |>
          filter(Pos != "") |>
          select(-any_of(c("Fav", "Share", "Cat Pos(Gen Pos)")))
        return(data)
      }
      
      colnames(data1) <- data1[1,]
      data1 <- data1[-1,]
      #If there is not distance in col names
      if(!("Dist" %in% names(data1))) {
        #Create a new column called distance
        data1$Dist <- get_distance(race)
      }
      
      #Only rbind with common columns
      common <- intersect(colnames(data), colnames(data1))
      data <- rbind(data[common], data1[common])
    }
    #Iterate through the links in the race
    for(link in page_links) {
      if(!is.na(link)) {
        url2 <- paste(base_address, link, sep="")
        html2 <- read_html(url2)
        
        tables2 <- html2 %>% html_table(fill=TRUE)
        data2 <- tables2[[2]]
        colnames(data2) <- data2[1,]
        data2 <- data2[-1,]
        
        #If there is not distance in col names
        if(!("Dist" %in% names(data2))) {
          #Create a new column called distance
          data2$Dist <- get_distance(race_names[[i]])
        }
        
        #Only rbind with common columns
        common <- intersect(colnames(data), colnames(data2))
        data <- rbind(data[common], data2[common])
      }
    }
    first_iteration <- 0
    i <- i+1
  }
  
  #Filter not needed columns
  data <- data[, !duplicated(colnames(data))]
  
  # Only include all ones where the person finished
  data <- data |>
    filter(Pos != "") |>
    select(!c("Fav", "Share", "Cat Pos(Gen Pos)"))
  
  
  return(data)
}
```

Now I can use these functions in order to gather the tables from the links:

```{r}
center_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1441")
sugar_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1446")
ledges_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1478")
summer_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1492")
yellow_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1575")
macbride_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1593&EId=2")
annett_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1546&EId=1")
jester_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1582")
```

Combine into one large data set???

...

Issues I had, and the solutions I produced:

...

### Cleaning the data

Add in additional cleaning data if needed

...


### Basic summaries

Include basic summaries if needed

...

### Research Questions Solutions

(Questions copied from Final Project Idea)

1)  Which cities/regions are most represented, and do runners from closer locations perform differently than those who travel farther?

...

2)  How does age group affect pace, and does that relationship change across different race distances?

...

3)  How does the performance gap between male and female athletes (measured by fastest race times) vary across age categories?

...

4)  Is there significant a relationship between race distance and the number of participants in running events?

...