---
title: "Final-Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyverse)
library(rvest)
library(lubridate)
library(RColorBrewer)
#install.packages("tidygeocoder")
library(tidygeocoder)

#install.packages("geosphere")
library(geosphere)
```

# DS2020: Final Project

Members: Samara Feldhacker (samara28), Alina Kirkpatrick (alinayjk)

### Data Gathering

In order to gather the data from online resources, I created functions that would take the base url and gather all of the information from the website that was needed. (More text yapping)

Creating functions:

```{r}
base_address <- "https://results.truetimeracing.com/"

get_distance <- function(race_name) {
  if(race_name == "4 Mile Run") {
    return(4.0)
  }
  if(race_name == "8 Mile Run") {
    return(8.0)
  }
  if (race_name == "Half Marathon") {
    return(13.1)
  }
  if (race_name == "10k Run") {
    return(6.2)
  }
  if (race_name == "5k Run") {
    return(3.1)
  }
  return(0.0)
}

information_scrapper <- function(url) {
  #First open the base url
  site <- read_html(url)
  
  #First gather all of the races
  race_links <- site %>% html_elements("#ctl00_Content_Main_divEvents .nav-link") %>% html_attr(name="href")
  race_names <- site %>% html_elements("#ctl00_Content_Main_divEvents .nav-link") %>% html_text()
  
  #Get the other page links for that race
  page_links <- site %>% html_elements("#ctl00_Content_Main_grdTopPager a") %>% html_attr(name="href")
  
  #Get the first table
  tables <- site %>% html_table(fill=TRUE)
  data <- tables[[2]]
  colnames(data) <- data[1,]
  data <- data[-1,]
  
  #If there is not distance in col names
  if(!("Dist" %in% names(data))) {
    #Create a new column called distance
    data$Dist <- get_distance(race_names[[1]])
  }
  
  first_iteration = 1
  i <- 1
  
  #Iterate through all of the races
  for(race in race_links) {
    if(!first_iteration) {
      #Read that specific race page
      url <- paste(base_address, race, sep="")
      html <- read_html(url)
      
      #Get the other page links for that race
      page_links <- html %>% html_elements("#ctl00_Content_Main_grdTopPager a") %>% html_attr(name="href")
      
      #Get the first table
      tables <- html %>% html_table(fill=TRUE)
      
      data1 <- tryCatch(tables[[2]], error = function(e) NULL)
      
      if(is.null(data1)) {
        data <- data[, !duplicated(colnames(data))]
        data <- data |>
          filter(Pos != "") |>
          select(-any_of(c("Fav", "Share", "Cat Pos(Gen Pos)")))
        return(data)
      }
      
      colnames(data1) <- data1[1,]
      data1 <- data1[-1,]
      #If there is not distance in col names
      if(!("Dist" %in% names(data1))) {
        #Create a new column called distance
        data1$Dist <- get_distance(race_names[[i]])
      }
      
      #Only rbind with common columns
      common <- intersect(colnames(data), colnames(data1))
      data <- rbind(data[common], data1[common])
    }
    #Iterate through the links in the race
    for(link in page_links) {
      if(!is.na(link)) {
        url2 <- paste(base_address, link, sep="")
        html2 <- read_html(url2)
        
        tables2 <- html2 %>% html_table(fill=TRUE)
        data2 <- tables2[[2]]
        colnames(data2) <- data2[1,]
        data2 <- data2[-1,]
        
        #If there is not distance in col names
        if(!("Dist" %in% names(data2))) {
          #Create a new column called distance
          data2$Dist <- get_distance(race_names[[i]])
        }
        
        #Only rbind with common columns
        common <- intersect(colnames(data), colnames(data2))
        data <- rbind(data[common], data2[common])
      }
    }
    first_iteration <- 0
    i <- i+1
  }
  
  #Filter not needed columns
  data <- data[, !duplicated(colnames(data))]
  
  # Only include all ones where the person finished
  data <- data |>
    filter(Pos != "") |>
    select(!c("Fav", "Share", "Cat Pos(Gen Pos)"))
  
  
  return(data)
}
```

Now I can use these functions in order to gather the tables from the links:

```{r}
center_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1441")
sugar_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1446")
ledges_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1478")
summer_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1492")
yellow_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1575")
macbride_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1593&EId=2")
annett_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1546&EId=1")
jester_trails <- information_scrapper("https://results.truetimeracing.com/results.aspx?CId=16535&RId=1582")
```

These will be the base data sets that will later get cleaned and combined to make them easier to work with.

...

Issues I had, and the solutions I produced:

...

### Cleaning the data

Now, using the tables produced, I decided to also create one large data set containing all of the races during the 2025 year. I am deciding to match the large data set with the same number of variables as the Center trails and sugar trails data set, but I am choosing to remove Race No, Finish (same as time), and Pace (which can be manual calculated in a much better format).

Step 1: Ensure all dataframes have the same number/type of variables:

```{r}
#Center and sugar are good

#Ledges is missing City, State, Pace, Start, Finish
ledges_trails <- ledges_trails |>
  mutate(City = "", State = "", Pace = "", Start = "", Finish = "")

#Summer is missing City, State, Pace
summer_trails <- summer_trails |>
  mutate(City = "", State = "", Pace = "")

#Yellow trails is missing city, State
yellow_trails <- yellow_trails |>
  mutate(City = "", State = "")

#Macbride has extra Age variable
macbride_trails <- macbride_trails |>
  select(!c("Age"))

#Annett is missing city, State, Pace
annett_trails <- annett_trails |>
  mutate(City = "", State = "", Pace = "")

#Jester is missing city, State
jester_trails <- jester_trails |>
  mutate(City = "", State = "")
```

Step 2: Clean the individual data sets, because I wanted to use another function.

```{r}
data_cleaner <- function(data, trailName) {
  #Remove Race No, Pace, Finish
  dataNew <- data |>
    select(!c("Race No", "Pace", "Finish"))
  
  #Turn gender into a factor
  dataNew$Gender <- factor(dataNew$Gender)
  
  #Turn Pos, Cat Pos, Gen pos, and Dist into a numerical variable
  dataNew$Pos <- as.numeric(dataNew$Pos)
  dataNew$`Cat Pos` <- as.numeric(dataNew$`Cat Pos`)
  dataNew$`Gen Pos` <- as.numeric(dataNew$`Gen Pos`)
  dataNew$Dist <- as.numeric(dataNew$Dist)
  
  #convert time, and start time into the time variable type
  dataNew$Start <- hms(dataNew$Start)
  dataNew$Time <- hms(dataNew$Time)
  
  #Turn category into a factor
  dataNew$Category <- factor(dataNew$Category)
  
  #Turn city and State into a factor - note may need to change later but whatever
  dataNew$City <- factor(dataNew$City)
  dataNew$State <- factor(dataNew$State)
  
  #Add on column for the trailName, and set as factor
  dataNew$Trail <- trailName
  
  dataNew$Trail <- factor(dataNew$Trail)
  
  return(dataNew)
}
```

```{r}
center_trails <- data_cleaner(center_trails, "Center")
sugar_trails <- data_cleaner(sugar_trails, "Sugar")
ledges_trails <- data_cleaner(ledges_trails, "Ledges")
summer_trails <- data_cleaner(summer_trails, "Summer")
yellow_trails <- data_cleaner(yellow_trails, "Yellow")
macbride_trails <- data_cleaner(macbride_trails, "MacBride")
annett_trails <- data_cleaner(annett_trails, "Annett")
jester_trails <- data_cleaner(jester_trails, "Jester")
```

Now, the dataframes will easily combine together into 1

```{r}
trail_races <- rbind(center_trails, sugar_trails, ledges_trails, summer_trails, yellow_trails, macbride_trails, annett_trails, jester_trails)

#Check the levels of Category, and manually rename to make consistent
levels(trail_races$Category)

levels(trail_races$Category) <- c("0-14", "20-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "Overall Winners", "15-19", "Unknown", "15-19", "20-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70+", "70+")

#reorder of age groups
levels(trail_races$Category)

trail_races$Category <- factor(trail_races$Category, levels = c("0-14", "15-19", "20-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70+", "Overall Winners", "Unknown"))

levels(trail_races$Category)
```
Extra data cleaning that is required for question 1:
Some of the cities are all caps while some have just the first letter caps, so I will convert these into their corresponding city. There are also a few cities that have some additional flaws to them. I am choosing to do this step by hand instead of creating a new function. It would be ideal for this to be a function, but some of the errors are not predictable.

```{r}
# Before
levels(trail_races$City)

# Manual change the levels where needed
levels(trail_races$City)[levels(trail_races$City) == "AMES"] <- "Ames"
levels(trail_races$City)[levels(trail_races$City) == "JeffersonJefferson"] <- "Jefferson"
levels(trail_races$City)[levels(trail_races$City) == "POLK CITY"] <- "Polk City"
levels(trail_races$City)[levels(trail_races$City) == "Waukeeukee"] <- "Waukee"
levels(trail_races$City)[levels(trail_races$City) == "CEDAR RAPIDS"] <- "Cedar Rapids"
levels(trail_races$City)[levels(trail_races$City) == "IOWA CITY"] <- "Iowa City"
levels(trail_races$City)[levels(trail_races$City) == "Iowa CityI"] <- "Iowa City"
levels(trail_races$City)[levels(trail_races$City) == "OttumwaOttumwa"] <- "Ottumwa"
levels(trail_races$City)[levels(trail_races$City) == "WATERLOO"] <- "Waterloo"


# After
levels(trail_races$City)
```

### Basic summaries

Include basic summaries if needed

```{r}
summary(trail_races)
```

### Research Questions Solutions

(Questions copied from Final Project Idea)

1)  Which cities/states are most represented, and do runners from closer locations perform differently than those who travel farther?

When answering the first part of the question, which cities/states are most represented, a graph that would easily show the most popular cities/states would be a bar chart. In this case, I will plot the top 20 most popular cities first.

```{r}
# Making a data set a copy of the original for this question, and remove not needed variables
trail_races_q1 <- trail_races |>
  select(!c("Name", "Pos", "Cat Pos", "Gen Pos", "Start"))

trail_races_q1 |>
  filter(!is.na(City), City != "") |>
  count(City, sort = TRUE) |>
  slice_head(n = 20) |>
  mutate(City = reorder(City, n)) |>
  ggplot(aes(x = n, y = City, fill = factor(n))) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.2) +
  labs(title = "Top 20 Cities by Trail Race Participation",
       x = "Number of participants",
       y = NULL) +
  scale_colour_discrete() +
  theme_bw()
  
```

As shown in the graph above, the majority of people are from larger population centers in Iowa, including Iowa City, Cedar Rapids, and Des Moines.

Now I will do the same thing, but for the states instead.

```{r}
trail_races_q1 |>
  filter(!is.na(State), State != "") |>
  count(State, sort = TRUE) |>
  mutate(State = reorder(State, n)) |>
  ggplot(aes(x = n, y = State, fill = factor(n))) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.2) +
  labs(title = "States by Trail Race Participation",
       x = "Number of participants",
       y = NULL) +
  scale_colour_discrete() +
  theme_bw()
```

As shown in the graph above, it is very obvious that the majority of people are from Iowa, or surrounding areas within the midwest.

Now when answering the second part of the question, we have to first determine a performance factor. In this case, time would not be a good factor because of the different race distances, and instead deriving pace in min/mile, would be an easy way to compare performance between groups. So, I am first going to calculate the pace using the existing Time and Dist variables.

```{r}
trail_races_q1 <- trail_races_q1 |>
  mutate(time_sec = as.numeric(Time),
         pace_min_mile = (time_sec / 60) / Dist)
```


runners from closer locations perform differently than those who travel farther?

```{r}
#First remove any null city/state values
trail_races_q1 <- trail_races_q1 |>
  filter(!is.na(State), State != "", !is.na(City), City != "")

# get the state abbreviation for each State
trail_races_q1 <- trail_races_q1 |>
  mutate(State_abbr = state.abb[match(State, state.name)])

# Combine city and state into 1 address
trail_races_q1 <- trail_races_q1 |>
  mutate(full_address = paste(City, State_abbr, sep = ", "))

# Get the lat and long from the address
trail_races_q1 <- trail_races_q1 |>
  geocode(address = full_address, method = "osm", lat = latitude, long = longitude)

# Now, I'm going to get the location of the races, note: I am manually doing this by looking up the venue locations and the corresponding lat/long. I could also make this a function, but doing it by hand would be easier in this case.

race_locations <- data.frame(
  Trail = c("Center", "Sugar", "Ledges", "Summer", "Yellow", "MacBride", "Annett", "Jester"),
  race_latitude = c(41.58687, 41.80723, 41.87665, 42.02676, 43.20060, 41.80723, 41.3605, 41.76110),
  race_longitude = c(-93.62495, -91.49406, -93.82328, -93.61704, -91.15254, -91.49406, -93.56133, -93.82439)
)

# Calculate distance from the race location

# Combine race_locations data and the trail_races data
trail_races_q1 <- trail_races_q1 |>
  left_join(race_locations, by = "Trail") |>
  rowwise() |>
  mutate(dist_to_race_miles = round(distHaversine(c(race_longitude,race_latitude), c(longitude, latitude)) * 0.00062137), 2) |>
  ungroup()
```
```{r}
# Everyone

trail_races_q1 |>
  ggplot(aes(x = dist_to_race_miles, y = pace_min_mile)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth() +
  scale_colour_brewer(type = "qual", palette = "Set1") +
  labs(title = "Travel Distance vs. Pace",
       x = "Distance Travelled to Race (miles)",
       y = "Pace (min/mile)") +
  theme_bw()

trail_races_q1 |>
  filter(dist_to_race_miles <= 500) |>
  ggplot(aes(x = dist_to_race_miles, y = pace_min_mile)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth() +
  scale_colour_brewer(type = "qual", palette = "Set1") +
  labs(title = "Travel Distance vs. Pace",
       x = "Distance Travelled to Race (miles)",
       y = "Pace (min/mile)") +
  theme_bw()
```
Based on gender:

```{r}
#display the data

trail_races_q1 |>
  filter(Dist <= 4, dist_to_race_miles <= 500) |>
  ggplot(aes(x = dist_to_race_miles, y = pace_min_mile, color = Gender)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth() +
  scale_colour_brewer(type = "qual", palette = "Set1") +
  labs(title = "Travel Distance vs. Pace for races 4 miles or less",
       x = "Distance Travelled to Race (miles)",
       y = "Pace (min/mile)", color = "Gender") +
  theme_bw()
```

```{r}
trail_races_q1 |>
  filter(Dist > 4, dist_to_race_miles <= 500) |>
  ggplot(aes(x = dist_to_race_miles, y = pace_min_mile, color = Gender)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth() +
  scale_colour_brewer(type = "qual", palette = "Set1") +
  labs(title = "Travel Distance vs. Pace for races over 4 miles",
       x = "Distance Travelled to Race (miles)",
       y = "Pace (min/mile)", color = "Gender") +
  theme_bw()

```


Based on age category:

```{r}
#display the data
trail_races_q1 |>
  filter(dist_to_race_miles <= 200) |>
  ggplot(aes(x = dist_to_race_miles, y = pace_min_mile, color = Category)) +
  geom_point(alpha = 0.2, size = 1.5) +
  geom_smooth(se = FALSE, method = "loess", span = 0.75) + 
  labs(title = "Travel Distance vs. Pace",
       x = "Distance Travelled to Race (miles)",
       y = "Pace (min/mile)", color = "Age Category") +
  theme_bw()
```

2)  How does age group affect pace, and does that relationship change across different race distances?

First I will recalculate pace into a new dataset for question 2, similar to how it was done for question 1. In addition, I will also remove not needed variables.

```{r}
trail_races_q2 <- trail_races |>
  mutate(time_sec = as.numeric(Time),
         pace_min_mile = (time_sec / 60) / Dist) |>
  select(!c("Name", "Pos", "Cat Pos", "Gen Pos", "Start", "City", "State"))
```

Next I will filter out the Overall Winners within Category, since their age is not listed, and any unkown values will be excluded

```{r}
trail_races_q2 <- trail_races_q2 |>
  filter(!is.na(Category), !Category %in% c("Overall Winners","Unknown"))
```

Now I will create a basic plot between age group and pace...

```{r fig.height=14, fig.width=10}
trail_races_q2 |>
  ggplot(aes(x = Category, y = pace_min_mile, fill = Category)) +
  geom_boxplot(alpha = 0.65, outlier.alpha = 0.25, outlier.size = 1, show.legend = FALSE) +
  facet_wrap(~ Dist, scales = "free_y", ncol = 1) +
  labs(title = "Pace by age group across race distances",
       x = "Age group",
       y = "Pace (min/mile)") +
  theme_bw()
```

Within the data set, the different distances include 3.1, 3.4, 3.65, 4, 6.2, 6.8, 8, 13.1. I will choose to group these distances into 3 distinctive groups. Those being 3-4 miles, 6-8 miles and then 13.1 miles. I will do this in order to more accuraely comapare short vs mid vs long races. Ill create a new distance_category label that just refactors the original

```{r}
levels(factor(trail_races_q2$Dist))

trail_races_q2$Dist_category <- factor(trail_races_q2$Dist)

levels(trail_races_q2$Dist_category) <- c("Short (3-4 mi)", "Short (3-4 mi)", "Short (3-4 mi)", "Short (3-4 mi)", "Mid (6-8 mi)", "Mid (6-8 mi)", "Mid (6-8 mi)", "Long (13.1 mi)")

levels(trail_races_q2$Dist_category)
```

Now I will create a box plot showing the age group vs. pace, faceted by distance

```{r fig.height=10, fig.width=10}
trail_races_q2 |>
  ggplot(aes(x = Category, y = pace_min_mile, fill = Category)) +
  geom_boxplot(alpha = 0.65,outlier.alpha = 0.25,outlier.size = 1, show.legend = FALSE) +
  facet_wrap(~ Dist_category, scales = "free_y", ncol = 1) +
  labs(title = "Pace by age group across race distances",
       x = "Age group",
       y = "Pace (min/mile)") +
  theme_bw()
```

Now to visual a closer represenation of this data, utilizing a heatmap may be good

```{r}
heatmap <- trail_races_q2 |>
  group_by(Category, Dist_category) |>
  summarise(median_pace = median(pace_min_mile, na.rm = TRUE), n = n(), .groups = "drop") |>
  filter(!is.na(median_pace))

mid_pace <- median(heatmap$median_pace)

heatmap |>
  ggplot(aes(x = Dist_category, y = Category, fill = median_pace)) +
  geom_tile(colour= "white") +
  geom_text(aes(label = sprintf("%d:%02d", floor(median_pace),round((median_pace %% 1) * 60))), size = 3) +
  scale_fill_gradient2(low = "green",mid = "white",high = "red", midpoint = median(mid_pace), name = "Median pace\n(min/mile)") +
  labs(title = "Median pace heat map — age group x race distance", 
       subtitle = "Green = faster, red = slower",
       x = "Race distance", 
       y = "Age group") +
  theme_bw()
```


3)  How does the performance gap between male and female athletes (measured by fastest race times) vary across age categories?

For easier time comparrisons, I made a new column where the time is mutated to be in seconds:
```{r}
Q3_df <- trail_races %>%
  mutate(time_seconds = as.numeric(Time))

Q3_df
```

Now, I will remove the 0.0 distance:
```{r}
Q3_df <- Q3_df %>%
  filter(!is.na(Dist), Dist != 0.0)

unique(Q3_df$Dist) #Checking it was removed (it was)
```
I will also create a new column that removes the category variable "Overall Winners" for simplicity:
```{r}
Q3_df <- Q3_df %>%
  filter(Category != "Overall Winners")

Q3_df # checking it worked
```
Because there are inconsistencies in some of the age catergories across race distances, we will only be removing the following age categories:
0-14, 15-19, 65-69, 70+
```{r}
Q3_df <- Q3_df %>%
  filter(!Category %in% c("0-14", "15-19", "65-69", "70+"))

unique(Q3_df$Category)

```

I will now find the top 3 runners for each trail, for each distance on the trail, based on gender, and age:

```{r}
top3_runners <- Q3_df %>%
  group_by(Trail, Dist, Category, Gender) %>%
  arrange(time_seconds, .by_group = TRUE) %>%
  mutate(rank_in_group = row_number()) %>%
  filter(rank_in_group <= 3)

top3_runners <- top3_runners %>%
  select(
    Trail,
    Dist, 
    Category,
    Gender,
    Name,
    time_seconds,
    rank_in_group
  )

top3_runners
# Making sure it made a new df instead of altering the original
trail_races
```

I am now going to summarize the average times for each age group and gender per race and distance:

```{r}
gap_run_time <- Q3_df %>%
  group_by(Trail, Dist, Category, Gender) %>%
  summarise(
    avg_time = mean(time_seconds, na.rm = TRUE),
    .groups = "drop"
  )

gap_run_time
```
From here, I will create a grouped bar-chart to visualize the data:
```{r}
#for better labeling
gap_run_time$Dist_label <- paste0("Dist - ", gap_run_time$Dist)


#the actual graph
ggplot(gap_run_time, aes(x = Category, y = avg_time, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  
  scale_fill_manual(values = c("Female" = "pink",
                               "Male" = "lightblue")) +
  
  facet_wrap(~Dist_label) +
  
  labs(
    title = "Performance Gap Between Male and Female Runners",
    x = "Age Category",
    y = "Average Time (seconds)"
  ) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )

```

4)  Is there significant a relationship between race distance and the number of participants in running events?

...